Indicators for Ecosystem Condition in Norway

### 2023-10-04 NDVI_seminat.Rmd
adding fig captions and caching some data

```{r where-are-seminat-sites, }
tm_shape(regnor) +
  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
  tm_borders() +
  tm_shape(SentinelNDVI.seminat) +
  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale = 2.5, legend.show = FALSE) +
  tm_add_legend(
    type = "symbol",
    col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
    labels = c(
      "T31 - boreal heath",
      "T32 - semi-natural grasslands",
      "T33 - semi-natural coastal grasslands",
      "T34 - coastal heathlands",
      "T40/41 - previously heavily altered grassland"
    ),
    title = "Major ecosystem type",
    size = 3
  ) +
  tm_layout(
    main.title = "Semi-natural ecosystems (NiN, Miljødirektoratets instruks)",
    main.title.size = 1.2, legend.position = c("right", "bottom"),
    legend.text.size = 1.3, legend.title.size = 1.4
  )
```

@@

```{r}
summary(SentinelNDVI.seminat$area_meters)
hist(SentinelNDVI.seminat$area_meters, xlim = c(0, 20000), breaks = 50000)
abline(v = 100, lty = 2)
```

semi-natural polygons stretch a large number of sizes from just above
100 sqm to almost 3 million sqm. 75% of all polygons are below 15297
sqm.

@@

```{r}
SentinelNDVI.seminat %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_grid(tilstand ~ hovedtype)
```

2022 stands out with the highest NDVI values missing. WE omit 2022 for
this analysis

@@

```{r warning=FALSE}
SentinelNDVI.seminat %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = hovedtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey")

@@

```{r warning=FALSE}
SentinelNDVI.seminat %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = region, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
```

@@

```{r warning=FALSE}
SentinelNDVI.seminat %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

@@

```{r warning=FALSE}
nin.seminat2 %>%
  filter(hovedtype == "T31") %>%
  filter(NiN_variable_code %in% c("7JB-BT", "7RA-BA", "7FA", "7TK")) %>%
  mutate(NiN_variable_des = recode(NiN_variable_code,
    "7JB-BT" = "Grazing pressure",
    "7RA-BA" = "Succession",
    "7FA" = "Alien species",
    "7TK" = "Heavy vehicles"
  )) %>%
  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
  geom_point(size = 2) +
  geom_jitter(width = 0.25, height = 0.25) +
  facet_wrap(~NiN_variable_des, nrow = 2, ncol = 3) +
  theme(legend.position = "none") +
  ggtitle("Condition variables in boreal heath (T31)") +
  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")

@@

```{r warning=FALSE, message=FALSE}
# NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
SentinelNDVI.seminat %>%
  group_by(id, year) %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_grid(region ~ hovedtype)
```

@@

```{r}
SentinelNDVI.seminat$mean_beta <- (SentinelNDVI.seminat$mean + 1) / 2
# NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
# we drop T41 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
SentinelNDVI.seminat.train <- SentinelNDVI.seminat %>%
  filter(year == kartleggingsaar) %>%
  filter(hovedtype != "T41")

@@

# We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
model.seminat.cond.Sent <- betareg(mean_beta ~ tilstand_num * region * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region != "Northern Norway", ])
model.seminat.cond.Sent <- StepBeta(model.seminat.cond.Sent)
summary(model.seminat.cond.Sent)
```

>>>>

```{r where-are-seminat-sites, fig.cap="A map of Norway showing the location of the semi-natural ecosystem sites."}
tm_shape(regnor) +
  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
  tm_borders() +
  tm_shape(SentinelNDVI.seminat) +
  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale = 2.5, legend.show = FALSE) +
  tm_add_legend(
    type = "symbol",
    col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
    labels = c(
      "T31 - boreal heath",
      "T32 - semi-natural grasslands",
      "T33 - semi-natural coastal grasslands",
      "T34 - coastal heathlands",
      "T40/41 - previously heavily altered grassland"
    ),
    title = "Major ecosystem type",
    size = 3
  ) +
  tm_layout(
    main.title = "Semi-natural ecosystems (NiN, Miljødirektoratets instruks)",
    main.title.size = 1.2, legend.position = c("right", "bottom"),
    legend.text.size = 1.3, legend.title.size = 1.4
  )
```

@@

```{r hist-sentinel-ndvis, fig.cap="A histogram showing the distribution of polygon sizes."}
summary(SentinelNDVI.seminat$area_meters)
hist(SentinelNDVI.seminat$area_meters, xlim = c(0, 20000), breaks = 50000)
abline(v = 100, lty = 2)
```

Semi-natural polygons stretch a large number of sizes from just above
100 sqm to almost 3 million sqm. 75% of all polygons are below 15297
sqm.

@@

```{r NDVI-values-over-time-ndvis, fig.cap="A figure showing NDVI values for semi-natural ecosystem over time for six different main nature types in varying degrees of condition."}
SentinelNDVI.seminat %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_grid(tilstand ~ hovedtype)
```

2022 stands out with the highest NDVI values missing. 
We omit 2022 for this analysis.

@@

```{r NDVI-major-basic-ndvis, warning=FALSE, fig.cap="NDVI values in major and basic semi natural ecosystem types in good condition."}
SentinelNDVI.seminat %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = hovedtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey")

@@

```{r ndvi-regions-ndvis, warning=FALSE, fig.cap="DVI variation across regions in semi-natural ecosystems (major types) in good condition"}
SentinelNDVI.seminat %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = region, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
```

@@

```{r warning=FALSE}
SentinelNDVI.seminat %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  theme(axis.text.x = element_text(angle = 35, vjust = 1, hjust=1))+
  facet_wrap(~hovedtype)
```

@@

```{r important-variables-ndvis, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?"}
nin.seminat2 %>%
  filter(hovedtype == "T31") %>%
  filter(NiN_variable_code %in% c("7JB-BT", "7RA-BA", "7FA", "7TK")) %>%
  mutate(NiN_variable_des = recode(NiN_variable_code,
    "7JB-BT" = "Grazing pressure",
    "7RA-BA" = "Succession",
    "7FA" = "Alien species",
    "7TK" = "Heavy vehicles"
  )) %>%
  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
  geom_point(size = 2) +
  geom_jitter(width = 0.25, height = 0.25) +
  facet_wrap(~NiN_variable_des, nrow = 2, ncol = 3) +
  theme(legend.position = "none") +
  ggtitle("Condition variables in boreal heath (T31)") +
  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")

@@

```{r balanced-data-ndvis, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradien, faceted by six semi natural ecosystem types in five regions in Norway. "}
# NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
SentinelNDVI.seminat %>%
  group_by(id, year) %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_grid(region ~ hovedtype)
```

@@

<!-- RUNTIME approx. 2 min -->
```{r, eval=F}
SentinelNDVI.seminat$mean_beta <- (SentinelNDVI.seminat$mean + 1) / 2
# NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
# we drop T41 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
SentinelNDVI.seminat.train <- SentinelNDVI.seminat %>%
  filter(year == kartleggingsaar) %>%
  filter(hovedtype != "T41")

@@

# We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
model.seminat.cond.Sent <- betareg(mean_beta ~ tilstand_num * region * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region != "Northern Norway", ])
model.seminat.cond.Sent <- StepBeta(model.seminat.cond.Sent)
summary(model.seminat.cond.Sent)

# Caching
saveRDS(model.seminat.cond.Sent, 
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.cond.Sent.rds")
```

```{r, include=}
model.seminat.cond.Sent <- readRDS( 
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.cond.Sent.rds")
summary(model.seminat.cond.Sent)
```
